### 2022-05-13 introduction.tex
Small fixes
Finished thesis. Implemented final feedback from instructor.

Solar power production is highly dependent on weather conditions \cite{lin_temporal_2020, lee_forecasting_2018, jaidee_very_2019, su_machine_2019, jang_solar_2016, jimenez_wrf-solar_2016, rognvaldsson_numerical_2013}. A single cloud can adversely impact the production of a solar power station. Such unreliability in a power production system is undesirable since production must match demand \cite{lee_forecasting_2018}. When solar power production falls, the power demand it was serving must be fulfilled by other means. Fossil fueled power stations are generally kept on standby, waiting to fulfil this demand \cite{lee_forecasting_2018}. These power stations often take hours to start, generally pollute more than other power stations and are expensive to run \cite{lee_forecasting_2018}.\\
Reliable and accurate forecasts of the fluctuations in solar power production can give power grid operators notice and insight into future power production \cite{lee_forecasting_2018}. This notice enables planning for the dips in power production and making arrangements for the future. This provides the tools to reduce the cost and pollution of power generation as well as increase the reliability of power delivery to consumers \cite{lee_forecasting_2018}.\\

@@

Here we explore using high-resolution weather forecast data from meteorological models, offloading a majority of this difficult task to the field of meteorology. Instead of generating forecasts, we focus on correlating existing forecasts with measurements and detecting and reducing temporal and spatial error in the general forecasts using newer methods like Temporal Fusion Transformers.
What is the quality of the predictions this methodology produces? Can we create a model which is useful to power producers in operating their systems more efficiently?\\

This thesis is structured as follows. First, we talked about the motivations for the work. Next, we discuss the background the work stands on. After that, we cover the methods utilized, the data and tools used to achieve these goals. The proceeding chapter contains the results of the work, and finally, we discuss the implications, strengths and flaws of the work, future work and end with conclusions.

>>>>

Solar power production is highly dependent on weather conditions \cite{lin_temporal_2020, lee_forecasting_2018, jaidee_very_2019, su_machine_2019, jang_solar_2016, jimenez_wrf-solar_2016, rognvaldsson_numerical_2013}. A single cloud can adversely impact the production of a solar power station. Such unreliability in a power production system is undesirable since production must match demand \cite{lee_forecasting_2018}. When solar power production falls, the power demand it was serving must be fulfilled by other means. Fossil fueled power stations are generally kept on standby, waiting to fulfil this demand \cite{lee_forecasting_2018}. These power stations often take hours to start, generally pollute more than other power stations, and are expensive to run \cite{lee_forecasting_2018}.\\
Reliable and accurate forecasts of the fluctuations in solar power production can give power grid operators notice and insight into future power production \cite{lee_forecasting_2018}. This notice enables planning for the dips in power production and making arrangements for the future. This provides the tools to reduce the cost and pollution of power generation as well as increase the reliability of power delivery to consumers \cite{lee_forecasting_2018}.\\

@@

Here we explore using high-resolution weather forecast data from meteorological models of irradiance, to improve the predictions  of solar power production. More specifically offloading a majority of this difficult task to the field of meteorology. Instead of generating forecasts, we focus on correlating existing forecasts with measurements and detecting and reducing temporal and spatial error in the general forecasts using newer methods like Temporal Fusion Transformers.
This thesis aims to address two questions: What is the quality of the predictions this methodology produces? Can we create a model which is useful to power producers in operating their systems more efficiently?\\

This thesis is structured as follows. First, we talk about the motivations for the work. Next, we discuss the background that the work stands on. After that, we cover the methods utilized, the data and tools used to achieve these goals. The following chapter contains the results of the work, and then we discuss the implications, strengths and weaknesses of the work. Finally, we conclude and discuss future work.

### 2022-05-13 methods.tex
Finished thesis. Implemented final feedback from instructor.

\subsection{Irradiance measurements}
Under certain circumstances irradiance values simulated after the fact can be accurate, however this does not apply in all conditions, and real measured values are much more reliable \cite{gueymard_global_2008}. The irradiance was measured with seven Kipp \& Zonen CM21 pyranometers placed around the power station. The pyranometers have a reported maximum measurement error of 3\%. For input into the network, the highest and lowest measured values are dropped and the average value of the other five measurements is used. This data is provided with five-minute intervals, but the neural network operates hourly. To downsample the measurement data to an hourly format, we averaged the values for the twenty minutes before and after the beginning of each hour. Finally, the irradiance values were normalized linearly to the range $0.0-1.0$ before being input into the neural network.

\subsection{Forecast Data\label{cha:forecast_data}}
The weather forecast data is provided by Belgingur ehf., Reykjavík, Iceland. We use irradiance values from historical 24-hour weather forecasts generated by their WRF model, discussed in Section~\ref{cha:WRF}. The forecasts have a temporal resolution of one hour and a spatial resolution of 2.5km. One of our models uses forecast data from a 3x3 grid centred on the solar power station, and the other model the data from only the centre of the grid. We use two irradiance values from the forecast, the downward irradiance and the downward diffuse irradiance. These irradiance values were normalized linearly to the range $0.0-1.0$ before being input into the neural network.

@@

    \noindent\textbf{Gating mechanisms} identify and skip unused input data, granting models greater ability to work with less curated input data, or datasets where it is unknown which parts are of significance.\\
    The Gated Residual Network (GRN) structure, shown in Figure~\ref{fig:ttgt_detail}, is a base building block of TFTs \cite{lim_temporal_2020}. It provides the model the flexibility to choose the complexity of its input, more complex where needed, and less complex where not.\\
    \clearpage
    \noindent\textbf{Variable selection networks} emphasize the most relevant inputs for every time step \cite{lim_temporal_2020}. Some variables may be more important than others, \enquote{TFT is designed to provide instance-wise variable selection through the use of variable selection networks applied to both static covariates and time-dependent covariates.} \cite{lim_temporal_2020} The variable selection mechanism also allows the model to remove noisy inputs that might have a negative effect on performance. \\
    \textbf{Static covariate encoders} encode static features into context vectors, integrating them into the network \cite{lim_temporal_2020}.  \\
    \textbf{Temporal processing} using a more traditional RNN sequence-to-sequence layer (see Section~\ref{cha:RNN}) for short term, local processing and the more novel multi-headed attention layer (see Section~\ref{cha:transformer}) for long term pattern capture \cite{lim_temporal_2020}.\\
    \textbf{Prediction intervals} in the form of quantile forecasts, discussed in Section~\ref{sec:loss_function}, display a confidence interval for the predictions \cite{lim_temporal_2020}. \\
    Figure~\ref{fig:ttgt_detail} shows a detailed view of the TFT structure as discussed.

    \afterpage{%
    \begin{figure}
        \centering
         \makebox[\textwidth][c]{\includegraphics[scale=0.36]{imgs/TFT.png}}
        \caption{A detailed view of the architecture of the Temporal Fusion transformer.\cite{lim_temporal_2020}
        \label{fig:ttgt_detail}}
    \end{figure}
    \clearpage
    }

@@

    \pagebreak
    \subsubsection{Tuning}
        Significant tuning was done to optimize the performance of both models. Since the goal of tuning was not only to minimize Mean Absolute Percentage Error but balance that with the ability to recognize uncertainty, the process was very manual. Tables~\ref{tab:full_params} and \ref{tab:lim_params} show the selected hyperparameters for the full and limited models respectively.\\

@@

    \pagebreak
    \subsubsection{Loss function}\label{sec:loss_function}
    The loss function used for Temporal Fusion Transformers is quantile loss, as it's an integral part of generating the confidence intervals of the model \cite{lim_temporal_2020}.
    A model's prediction is a random value within a probability distribution \cite{wen_multi-horizon_2018}. If we make multiple predictions we can estimate this probability distribution by seeing within what range a selected proportion of the predictions fall \cite{koenker_quantile_2001}. PyTorch Forecasting's quantile loss function takes the Mean Absolute Error (MAE) of three quantiles, 90\%, 50\% and 10\%. This gives a good picture of the model's confidence in a prediction. A smaller quantile range indicates more confidence.

>>>>

\subsection{Irradiance measurements}
Under certain circumstances irradiance values simulated after the fact can be accurate, however, this does not apply in all conditions, and real measured values are much more reliable \cite{gueymard_global_2008}. The irradiance was measured with seven Kipp \& Zonen CM21 pyranometers placed around the power station. The pyranometers have a reported maximum measurement error of 3\%. For input into the network, the highest and lowest measured values are dropped and the average value of the other five measurements is used. This data is provided with five-minute intervals, but the neural network operates hourly. To downsample the measurement data to an hourly format, we averaged the values for the twenty minutes before and after the beginning of each hour. Finally, the irradiance values were normalized linearly to the range $0.0-1.0$ before being input into the neural network.

\subsection{Forecast data\label{cha:forecast_data}}
The weather forecast data is provided by Belgingur ehf., Reykjavík, Iceland. We use irradiance values from historical 24-hour weather forecasts generated by their WRF model, discussed in Section~\ref{cha:WRF}. The forecasts have a temporal resolution of one hour and a spatial resolution of 2.5km. One of our models uses forecast data from a 3x3 grid centred on the solar power station, and the other model the data from only the centre of the grid. We use two irradiance values from the forecast, the downward irradiance and the downward diffuse irradiance. These irradiance values were also normalized linearly to the range $0.0-1.0$ before being input into the neural network.

@@

    \noindent\textbf{Gating mechanisms} identify and skip unused input data, granting models greater ability to work with less curated input data or datasets where it is unknown which parts are of significance.

    The Gated Residual Network (GRN) structure, shown in Figure~\ref{fig:ttgt_detail}, is a base building block of TFTs \cite{lim_temporal_2020}. It provides the model with the flexibility to choose the complexity of its input, more complex where needed, and less complex where not.

    \clearpage
    \noindent\textbf{Variable selection networks} emphasize the most relevant inputs for every time step \cite{lim_temporal_2020}. Some variables may be more important than others, \enquote{TFT is designed to provide instance-wise variable selection through the use of variable selection networks applied to both static covariates and time-dependent covariates.} \cite{lim_temporal_2020} The variable selection mechanism also allows the model to remove noisy inputs that might have a negative effect on performance. \\
    \textbf{Static covariate encoders} encode static features into context vectors, integrating them into the network \cite{lim_temporal_2020}.  \\
    \textbf{Temporal processing} using a more traditional RNN sequence-to-sequence layer (see Section~\ref{cha:RNN}) for short term, local processing and the more novel multi-headed attention layer (see Section~\ref{cha:transformer}) for long term pattern capture \cite{lim_temporal_2020}.\\
    \textbf{Prediction intervals} in the form of quantile forecasts, discussed in Section~\ref{sec:loss_function}, display a confidence interval for the predictions \cite{lim_temporal_2020}. \\
    Figure~\ref{fig:ttgt_detail} shows a detailed view of the TFT structure as discussed.

    \afterpage{%
    \begin{figure}
        \centering
         \makebox[\textwidth][c]{\includegraphics[scale=0.36]{imgs/TFT.png}}
        \caption{A detailed view of the architecture of the Temporal Fusion transformer. \cite{lim_temporal_2020}
        \label{fig:ttgt_detail}}
    \end{figure}
    \clearpage
    }

@@

    \pagebreak
    \subsubsection{Tuning}
        Significant tuning was done to optimize the performance of both models. Since the goal of tuning was not only to minimize Mean Absolute Percentage Error but balance that with the ability to recognize uncertainty, the process was very manual. Tables~\ref{tab:full_params} and \ref{tab:lim_params} show the selected hyperparameters for the full and limited models respectively.

@@

    \pagebreak
    \subsubsection{Loss function}\label{sec:loss_function}
    The loss function used for Temporal Fusion Transformers is quantile loss, as it is an integral part of generating the confidence intervals of the model \cite{lim_temporal_2020}.
    A model's prediction is a random value within a probability distribution \cite{wen_multi-horizon_2018}. If we make multiple predictions we can estimate this probability distribution by seeing within what range a selected proportion of the predictions fall \cite{koenker_quantile_2001}. PyTorch Forecasting's quantile loss function takes the Mean Absolute Error (MAE) of three quantiles, 90\%, 50\% and 10\%. This gives a good picture of the model's confidence in a prediction. A smaller quantile range indicates more confidence.

### 2022-05-13 results.tex
Small fixes
Finished thesis. Implemented final feedback from instructor.

\chapter{Results}
Both models show a strong ability to predict solar irradiance one day into the future.
The full model, using weather forecast data from the surrounding area, shows both a lower Mean Absolute Percentage Error (MAPE), at 13.21\% vs. 15.98\%, as well as better tracking in challenging conditions and a better ability to recognize uncertain conditions, generally reporting lower confidence when its predictions are wrong. Both models can be trained to achieve slightly lower MAPE, but that comes at the cost of a higher tendency to underfit variance and higher false confidence.
We show irradiance plots for select days representative of the models' performance in various conditions. These plots show the measured irradiance for the day, the models' predicted irradiance values, along with a 90\% confidence interval, detailed in section~\ref{sec:loss_function}. The plots also show the forecast downward shortwave irradiance, the largest factor in measured irradiance, from the model input. Additionally, the graphs contain the average MAPE for the day and the sum of prediction and observation values. The sum values act as a gauge for the total irradiance for the day.

@@

In Figure~\ref{fig:full_med_med} we see examples of days where the model achieves higher error, MAPE in the range 19-35\%, but with generally larger 90\% confidence intervals so the model still keeps the measured values close to or within them.
\begin{figure}[ht!]
    \centering
    \subfloat[]{\includegraphics[scale=0.5]{imgs/graphs/full/medium/med_m_1.png}}\qquad
    \subfloat[]{\includegraphics[scale=0.5]{imgs/graphs/full/medium/med_m_2.png}}\qquad
    \subfloat[]{\includegraphics[scale=0.5]{imgs/graphs/full/medium/med_m_3.png}}\qquad
    \subfloat[]{\includegraphics[scale=0.5]{imgs/graphs/full/medium/med_m_4.png}}\qquad
    \subfloat[]{\includegraphics[scale=0.5]{imgs/graphs/full/medium/med_m_5.png}}\qquad
    \caption{Examples of days with some drops in irradiance, where the model performs somewhat well.
    \label{fig:full_med_med}}
\end{figure}

For a few medium variance days, the model does not track well. Figures~\ref{fig:full_med_bad_a} and \ref{fig:full_med_bad_d} show where the model predicts a constant lower irradiance when half the day had high irradiance and the other half low. Figures~\ref{fig:full_med_bad_b} and \ref{fig:full_med_bad_c} show the model predicting a mostly or fully sunny day when it wasn't. In all of these cases, the observed irradiance is far outside the 90\% confidence interval.
\begin{figure}[ht!]
    \centering
    \subfloat[]{\includegraphics[scale=0.5]{imgs/graphs/full/medium/med_b_1.png}\label{fig:full_med_bad_a}}\qquad
    \subfloat[]{\includegraphics[scale=0.5]{imgs/graphs/full/medium/med_b_2.png}\label{fig:full_med_bad_b}}\qquad
    \subfloat[]{\includegraphics[scale=0.5]{imgs/graphs/full/medium/med_b_3.png}\label{fig:full_med_bad_c}}\qquad
    \subfloat[]{\includegraphics[scale=0.5]{imgs/graphs/full/medium/med_b_4.png}\label{fig:full_med_bad_d}}\qquad
    \caption{Examples of days with some drops in irradiance, where the model performs badly.
    \label{fig:full_med_bad}}
\end{figure}

>>>>

\chapter{Results}
Both models show a strong ability to predict solar irradiance one day into the future.
The full model, using weather forecast data from the surrounding area, shows both a lower Mean Absolute Percentage Error (MAPE), at 13.21\% vs. 15.98\%, as well as better tracking in challenging conditions and a better ability to recognize uncertain conditions, generally reporting lower confidence when its predictions are wrong. Both models can be trained to achieve slightly lower MAPE, but that comes at the cost of a higher tendency to underfit variance and higher false confidence.
We show irradiance plots for select days representative of the models' performance in various conditions. These plots show the measured irradiance for the day, the models' predicted irradiance values, along with a 90\% confidence interval, detailed in Section~\ref{sec:loss_function}. The plots also show the forecast downward shortwave irradiance, the largest factor in measured irradiance, from the model input. Additionally, the graphs contain the average MAPE for the day and the sum of prediction and observation values. The sum values act as a gauge for the total irradiance for the day.

@@

In Figure~\ref{fig:full_med_med} we see examples of days where the model achieves higher error, MAPE in the range 19-35\%, but with generally larger 90\% confidence intervals so the model still keeps the measured values close to or within them.

\afterpage{%
\begin{figure}
    \centering
    \subfloat[]{\includegraphics[scale=0.5]{imgs/graphs/full/medium/med_m_1.png}}\qquad
    \subfloat[]{\includegraphics[scale=0.5]{imgs/graphs/full/medium/med_m_2.png}}\qquad
    \subfloat[]{\includegraphics[scale=0.5]{imgs/graphs/full/medium/med_m_3.png}}\qquad
    \subfloat[]{\includegraphics[scale=0.5]{imgs/graphs/full/medium/med_m_4.png}}\qquad
    \subfloat[]{\includegraphics[scale=0.5]{imgs/graphs/full/medium/med_m_5.png}}\qquad
    \caption{Examples of days with some drops in irradiance, where the model performs somewhat well.
    \label{fig:full_med_med}}
\end{figure}
\clearpage
}

For a few medium variance days, the model does not track well. Figures~\ref{fig:full_med_bad_a} and \ref{fig:full_med_bad_d} show where the model predicts a constant lower irradiance when half the day had high irradiance and the other half low. Figures~\ref{fig:full_med_bad_b} and \ref{fig:full_med_bad_c} show the model predicting a mostly or fully sunny day when it was not. In all of these cases, the observed irradiance is far outside the 90\% confidence interval.

\afterpage{%
\begin{figure}
    \centering
    \subfloat[]{\includegraphics[scale=0.5]{imgs/graphs/full/medium/med_b_1.png}\label{fig:full_med_bad_a}}\qquad
    \subfloat[]{\includegraphics[scale=0.5]{imgs/graphs/full/medium/med_b_2.png}\label{fig:full_med_bad_b}}\qquad
    \subfloat[]{\includegraphics[scale=0.5]{imgs/graphs/full/medium/med_b_3.png}\label{fig:full_med_bad_c}}\qquad
    \subfloat[]{\includegraphics[scale=0.5]{imgs/graphs/full/medium/med_b_4.png}\label{fig:full_med_bad_d}}\qquad
    \caption{Examples of days with some drops in irradiance, where the model performs badly.
    \label{fig:full_med_bad}}
\end{figure}
\clearpage
}
